---
title: "Basics of Panel Data"
author: "Ian McCarthy | Emory University"
format: 
  revealjs:
    theme: [moon]
    preview-links: auto
    chalkboard:
      boardmarker-width: 5
    slide-number: true
    width: 1600
    height: 900    
#    embed-resources: true
from: markdown+emoji
execute: 
  echo: true
---

```{r}
#| include: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, lubridate, gapminder, scales, lmtest, DIDmultiplegt,
               hrbrthemes, gganimate, OECD, causaldata, fixest, modelsummary, plm)
```


# Panel Data and Fixed Effects


---

## Basics of panel data

- Repeated observations of the same units over time (balanced vs unbalanced)
- Identification due to variation **within unit**

::: {.fragment}
**Notation**

- Unit $i=1,...,N$ over several periods $t=1,...,T$, which we denote $y_{it}$
- Treatment status $D_{it}$
- Regression model, <br>
$y_{it} = \delta D_{it} + \gamma_{i} + \gamma_{t} + \epsilon_{it}$ for $t=1,...,T$ and $i=1,...,N$
:::

---

## Benefits of Panel Data

- *May* overcome certain forms of omitted variable bias
- Allows for unobserved but time-invariant factor, $\gamma_{i}$, that affects both treatment and outcomes

::: {.fragment}
**Still assumes**

- No time-varying confounders 
- Past outcomes do not directly affect current outcomes
- Past outcomes do not affect treatment (reverse causality)
:::

---

## Some textbook settings

- Unobserved "ability" when studying schooling and wages
- Unobserved "quality" when studying physicians or hospitals

---

## Fixed effects and regression

$y_{it} = \delta D_{it} + \gamma_{i} + \gamma_{t} + \epsilon_{it}$ for $t=1,...,T$ and $i=1,...,N$

::: {.fragment}

- Allows correlation between $\gamma_{i}$ and $D_{it}$
- Physically estimate $\gamma_{i}$ in some cases via set of dummy variables
- More generally, "remove" $\gamma_{i}$ via:
  - "within" estimator
  - first-difference estimator
:::

---

## Within Estimator

$y_{it} = \delta D_{it} + \gamma_{i} + \gamma_{t} + \epsilon_{it}$ for $t=1,...,T$ and $i=1,...,N$

::: {.fragment}

- Most common approach (default in most statistical software)
- Equivalent to demeaned model:
$$y_{it} - \bar{y}_{i} = \delta (D_{it} - \bar{D}_{i}) + (\gamma_{i} - \bar{\gamma}_{i}) + (\gamma_{t} - \bar{\gamma}_{t}) + (\epsilon_{it} - \bar{\epsilon}_{i})$$

- $\gamma_{i} - \bar{\gamma}_{i} = 0$ since $\gamma_{i}$ is time-invariant
- Requires *strict exogeneity* assumption (error is uncorrelated with $D_{it}$ for all time periods)
:::

---

## First-difference

$y_{it} = \delta D_{it} + \gamma_{i} + \gamma_{t} + \epsilon_{it}$ for $t=1,...,T$ and $i=1,...,N$

::: {.fragment}

- Instead of subtracting the mean, subtract the prior period values
$$y_{it} - y_{i,t-1} = \delta(D_{it} - D_{i,t-1}) + (\gamma_{i} - \gamma_{i}) + (\gamma_{t} - \gamma_{t-1}) + (\epsilon_{it} - \epsilon_{i,t-1})$$

- Requires exogeneity of $\epsilon_{it}$ and $D_{it}$ only for time $t$ and $t-1$ (weaker assumption than within estimator)
- Sometimes useful to estimate both FE and FD just as a check
:::

---

## Keep in mind...

- Discussion only applies to linear case or very specific nonlinear models
- Fixed effects at lower "levels" accommodate fixed effects at higher levels (e.g., FEs for hospital combine to form FEs for zip code, etc.)
- Fixed effects can't solve reverse causality
- Fixed effects don't address unobserved, time-varying confounders
- Can't estimate effects on time-invariant variables
- May "absorb" a lot of the variation for variables that don't change much over time


---

## Within Estimator (Default) in practice

```{r}
#| eval=FALSE

library(fixest)
library(causaldata)
reg.dat <- causaldata::gapminder %>%
  mutate(lgdp_pc=log(gdpPercap))
feols(lifeExp~lgdp_pc | country, data=reg.dat)
```


---

## Within Estimator (Default) in practice

```{r}
#| code-fold: true
#| code-summary: "R Code"

library(fixest)
library(modelsummary)
library(causaldata)
reg.dat <- causaldata::gapminder %>%
  mutate(lgdp_pc=log(gdpPercap))
m1 <- feols(lifeExp ~ lgdp_pc | country, data=reg.dat)
modelsummary(list("Default FE"=m1), 
             shape=term + statistic ~ model, 
             gof_map=NA, 
             coef_rename=c("lgdp_pc"="Log GDP per Capita"))
```


---

## Within Estimator (Manually Demean) in practice

```{r}
#| eval=FALSE

library(causaldata)
reg.dat <- causaldata::gapminder %>%
  mutate(lgdp_pc=log(gdpPercap)) %>%
  group_by(country) %>%
  mutate(demean_lifeexp=lifeExp - mean(lifeExp, na.rm=TRUE),
         demean_gdp=lgdp_pc - mean(lgdp_pc, na.rm=TRUE))
lm(demean_lifeexp~ 0 + demean_gdp, data=reg.dat)
```

---

## Within Estimator (Manually Demean) in practice
```{r}
#| code-fold: true
#| code-summary: "R Code"

library(lmtest)
reg.dat <- causaldata::gapminder %>%
  group_by(country) %>%
  mutate(lgdp_pc=log(gdpPercap),
         lgdp_pc=lgdp_pc - mean(lgdp_pc, na.rm=TRUE),
         lifeExp=lifeExp - mean(lifeExp, na.rm=TRUE))

m2 <- lm(lifeExp~ 0 + lgdp_pc , data=reg.dat)
modelsummary(list("Default FE"=m1, "Manual FE"=m2), 
             shape=term + statistic ~ model, 
             gof_map=NA, 
             coef_rename=c("lgdp_pc"="Log GDP per Capita"),
             vcov = ~country)
```

**Note:** `feols` defaults to clustering at level of FE, `lm` requires our input

---

## First differencing (default) in practice

```{r}
#| eval=FALSE

library(plm)
reg.dat <- causaldata::gapminder %>%
  mutate(lgdp_pc=log(gdpPercap))

plm(lifeExp ~ 0 + lgdp_pc, model="fd", individual="country", index=c("country","year"), data=reg.dat)
```

---

## First differencing (manual) in practice

```{r}
#| code-fold: true
#| code-summary: "R Code"

library(plm)
reg.dat <- causaldata::gapminder %>%
  mutate(lgdp_pc=log(gdpPercap))

m3 <- plm(lifeExp ~ 0 + lgdp_pc, model="fd", index=c("country","year"), data=reg.dat)

modelsummary(list("Default FE"=m1, "Manual FE"=m2, "Default FD"=m3), 
             shape=term + statistic ~ model, 
             gof_map=NA, 
             coef_rename=c("lgdp_pc"="Log GDP per Capita"))
```



---

## First differencing (manual) in practice

```{r}
#| eval=FALSE

reg.dat <- causaldata::gapminder %>%
  mutate(lgdp_pc=log(gdpPercap)) %>%  
  group_by(country) %>%
  arrange(country, year) %>%
  mutate(fd_lifeexp=lifeExp - lag(lifeExp),
         lgdp_pc=lgdp_pc - lag(lgdp_pc)) %>%
  na.omit()

lm(fd_lifeexp~ 0 + lgdp_pc , data=reg.dat)
```

---

## First differencing (manual) in practice

```{r}
#| code-fold: true
#| code-summary: "R Code"

reg.dat <- causaldata::gapminder %>%
  mutate(lgdp_pc=log(gdpPercap)) %>%  
  group_by(country) %>%
  arrange(country, year) %>%  
  mutate(fd_lifeexp=lifeExp - dplyr::lag(lifeExp),
         lgdp_pc=lgdp_pc - dplyr::lag(lgdp_pc)) %>%
  na.omit()

m4 <- lm(fd_lifeexp~ 0 + lgdp_pc , data=reg.dat)
modelsummary(list("Default FE"=m1, "Manual FE"=m2, "Default FD"=m3, "Manual FD"=m4), 
             shape=term + statistic ~ model, 
             gof_map=NA, 
             coef_rename=c("lgdp_pc"="Log GDP per Capita"))
```


---

## FE and FD with same time period

```{r}
#| code-fold: true
#| code-summary: "R Code"

reg.dat2 <- causaldata::gapminder %>%
  mutate(lgdp_pc=log(gdpPercap)) %>%
  inner_join(reg.dat %>% select(country, year), by=c("country","year"))
m5 <- feols(lifeExp ~ lgdp_pc | country, data=reg.dat2)
modelsummary(list("Default FE"=m5, "Default FD"=m3, "Manual FD"=m4), 
             shape=term + statistic ~ model, 
             gof_map=NA, 
             coef_rename=c("lgdp_pc"="Log GDP per Capita"))
```

Don't want to read too much into this, but...

- Likely strong serial correlation in this case (almost certainly)
- Mispecified model


